Social Participation Ontology: community documentation, enhancements and use examples

### 2015-01-10 paper.tex
terminado artigo, faltando apenas completar as referencias

\section{Upper and domain ontologies}\label{upper}
There are various ontologies of interest for participatory democracy. To exemplify, three upper ontologies were chosen:
\begin{itemize}
    \item BFO $\rightarrow$ the Basic Formal Ontology is oriented to scientific use of data, which is confluent with Brazilian hackers and scientific community~\cite{bfo}. 
    \item DBpedia $\rightarrow$ ``a crowd-sourced community effort to extract structured information from Wikipedia''. Convenient for bootstrapping the OPS to out-of-domain knowledge with crowd-sourced structured information~\cite{dbpedia}.
    \item W3C recommendations for open-government, these are the work product of specialized groups and are regarded in high account by the semantic web community~\cite{egovW3C}.
\end{itemize}

\section{Restrictions in \vcps\ and initial \ops\ which were removed for the time being}\label{ap:restr}

These restrictions were part of \vcps\ documentation, but not implemented in \vcps\ \owl\ code. These were implemented in preliminary \ops\ implementation, although removed from current \ops\ as explained in Section~\ref{impl}:
\begin{itemize}
    \item Restriction: Role hasParticipationCharacteristic some ParticipationCharacteristic.
    \item Restriction: Results composesSolution some Solution.
    \item Restriction: Problem generatesCause some Cause.
\end{itemize}
\begin{table}
  \centering
  \caption{Restrictions of the OSP (Ontology of Social Participation). All OSP restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & startsCause        & Cause\\\hline
Supporter    & supportsCause      & Cause\\\hline
Executer     & executesAction     & Action\\\hline
Solution     & solvesProblem      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & producesResults    & Results\\\hline
Result       & composesSolution   & Solution\\\hline
Cause        & dasAction          & Action\\\hline
Action       & belongsToScope     & Scope\\\hline
Cause        & hasTheme           & Theme\\\hline
Cause        & proposesSolution   & Solution\\\hline
Problem      & generatesCause     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

It is important to realize that OSP is dedicated to a complete social participation process, with a beginning, middle and end.
Thus, such OWL restrictions are valid for the final state of the process (for example, in an arbitrary snapshot, a \texttt{SocialActor} may be not tied to a role). This might lead to NULL or ``not yet defined'' field supplies.

Also, in \vcps, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction.

\section{Public consultation}\label{ap:public}
Public consultations have core importance in participatory democracy and the model in figure~\ref{fig:consult} was conceived as part of OSP community documentation.
Therefore, an OWL implementation of such public consultation, should be developed and made public, with dedicated attention.
Considerations of this public consultation with respect to the OSP and formal existent instances should be addressed as well.
This public consultation model is observed here for completeness of exposition and further details goes beyond the scope of this article.

\begin{figure*}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/publicConsultation__}
    \caption{A diagram representation of a general public consultation. The OWL implementation of this model is in subsection~\ref{public}.}
    \label{fig:consult}
\end{figure*}

>>>>

%
%\section{Upper and domain ontologies}\label{upper}
%There are various ontologies of interest for participatory democracy and thus for \ops. To exemplify, three sources were chosen:
%\begin{itemize}
%    \item DBpedia $\rightarrow$ ``a crowd-sourced community effort to extract structured information from Wikipedia''. Convenient for bootstrapping the OPS to out-of-domain knowledge with crowd-sourced structured information~\cite{dbpedia}.
%    \item W3C recommendations for open-government, these are the work product of specialized groups and are regarded in high account by the semantic web community~\cite{egovW3C}.
%    \item Current upper ontologies used in \ops: \foaf\ and \bfo.
%\end{itemize}
%
%This list is endless and further work should narrow the interest to essential upper ontologies. As part of the linkage to other participatory instances and even to other domains, this is a core 
%
\section{Restrictions in \vcps\ and initial \ops\ which were removed for the time being}\label{ap:restr}

These restrictions were part of \vcps\ documentation, but not implemented in \vcps\ \owl\ code:
\begin{itemize}
    \item Restriction: {\tt ops:Role ops:hasParticipationCharacteristic some ops:ParticipationCharacteristic}.
    \item Restriction: {\tt ops:Results ops:composesSolution some ops:Solution}.
    \item Restriction: {\tt Problem generatesCause some Cause}.
\end{itemize}

Present in preliminary \ops\ implementation, all restrictions were removed from current \ops\ as explained in Section~\ref{impl}
\begin{table}
  \centering
  \caption{Restrictions of the OSP (Ontology of Social Participation). All OSP restrictions are existential (\texttt{owl:someValuesFrom)}.}
  \begin{tabular}{|l|l|l|}\hline
{\bf subject} & {\bf predicate} & {\bf object} \\\hline\hline
Initiator    & startsCause        & Cause\\\hline
Supporter    & supportsCause      & Cause\\\hline
Executer     & executesAction     & Action\\\hline
Solution     & solvesProblem      & Problem\\\hline
SocialActor  & hasRole            & Role\\\hline
Action       & producesResults    & Results\\\hline
Result       & composesSolution   & Solution\\\hline
Cause        & dasAction          & Action\\\hline
Action       & belongsToScope     & Scope\\\hline
Cause        & hasTheme           & Theme\\\hline
Cause        & proposesSolution   & Solution\\\hline
Problem      & generatesCause     & Cause\\\hline
  \end{tabular}
  \label{ospRestr}
\end{table}

Such \owl\ restrictions are valid, to the best of chances, for the final state of a participatory process (for example, in an arbitrary snapshot, a \texttt{SocialActor} may be not tied to a role). This might lead to NULL or ``not yet defined'' field supplies.
Also, in \vcps, existential restrictions were  written as ``min 1''. 
These were changed to the standard ``some'' existential restriction in preliminary \ops\ and were completely removed afterwards, as a way to avoid making \ops\ usage rigid and a more complicated.

\section{Public consultation}\label{ap:public}
Public consultations have core importance in participatory democracy and the model in figure~\ref{fig:consult} was conceived as part of \vcps\ community documentation.
An \owl\ implementation of such public consultation was meant to be developed and made public, with dedicated attention.
Considerations of this public consultation with respect to the \ops\ and formal existent instances was envisioned as well.
This public consultation model is observed here for completeness of exposition and further details goes beyond the scope of this article. A public consultations along the lines of a recent Presidential decree~\cite{decree} was formalized in \obs\ \owl\ and in \vbs\ \skos~\cite{pnud5}.

\begin{figure*}[H]
    \centering
    \includegraphics[width=0.9\textwidth]{figs/publicConsultation__}
    \caption{A diagram representation of a general public consultation. Further context of this image is in Appendix~\ref{ap:public}.}
    \label{fig:consult}
\end{figure*}
